\vspace*{-8mm}\textbf{\textsc{Novelty Detection for Semantic Place Categorization}}

### 2011-07-04 thesis/abstract.tex
Some language fixes by LPR and Michael.

This thesis studies a semantic mapping process that employs spatial semantic knowledge
and represents it using a probabilistic graphical model. It develops methods 
for performing detection of novel room categories which were not previously known to the robot applicable 
to that graphical model. Finally, it draws attention to the usefulness of unlabelled data for the 
novelty detection process in order to considerably increase the detection accuracy.

@@

Já à muitos anos que a humanidade sonha que um dia robôs caminharam lado a lado com os humanos.
Acredita-se que eles serão capaz de se mover e interagir connosco, compreender os nossos conceitos
e serem capazes de pensar.
Um passo importante nesta direcção é a criação de robôs com conhecimento de conceitos e significados
próprios do mundo humano.
No entanto é irrealista acreditar que será possível descrever e representar em toda a sua extensão
tais conceitos. Assim os agentes deverão ser capazes de aprender e adaptar o seu conhecimento quando
confrontados com novas situações.
A detecção destas novas situações, onde o conhecimento de um agente não é suficiente, têm um papel importante
na adaptação e aprendizagem de novos conceitos e significados e é por isso o tópico principal desta tese.

>>>>

This thesis studies a semantic mapping process that employs spatial semantic knowledge
and represents it using a probabilistic graphical model. It develops methods 
for performing detection of novel room categories which were not previously known to the robot,
using for that graphical models. Finally, it draws attention to the usefulness of unlabelled data for the 
novelty detection process in order to considerably increase the detection accuracy.

@@

Já há muitos anos que a humanidade sonha que um dia robôs caminharam lado a lado com os humanos.
Acredita-se que eles serão capaz de se mover e interagir connosco, compreender os nossos conceitos
e serem capazes de pensar.
Um passo importante nesta direcção é a criação de robôs com conhecimento de conceitos e significados
próprios do mundo humano.
No entanto é irrealista acreditar que será possível descrever e representar em toda a sua extensão
tais conceitos. Assim os agentes deverão ser capazes de aprender e adaptar o seu conhecimento quando
confrontados com novas situações.
A detecção destas novas situações, onde o conhecimento de um agente não é suficiente, têm um papel importante
na adaptação e aprendizagem de novos conceitos e significados e é por isso o tópico principal desta tese.

### 2011-07-04 thesis/background.tex
Some language fixes by LPR and Michael.

For example \glspl{SVM}, in their basic form, are only able to handle linear spaces.
But the classes are most of the time not linearly separable in the input
space. Although there might exists a transformation $\phi$ from the original
space into a space $H$ where the input becomes linearly separable.

@@

The most used method was presented by \cite{lowe1999object}. And its based on
building a feature vector for each image. Each of those features is based on
\emph{interesting points} detected by detecting maxima and minima of a
difference of Gaussian functions applied in a scale-space.
The scale space is used to provide scale invariant detection. Gaussian functions
are used as they are the only way to model a linear scale-space.
Each interesting point is then described by a container that is rotation
invariant.

@@

\subsubsection*{Gist of a Scene}
\label{sec:gist}
\cite{oliva2006building} argue that fast scene recognition does not need to be
built on top of object recognition but can be analyzed by scene-centered
mechanisms.
They defend that position by pointing out behaviours on human vision:
when provided with a glance of a shot a person can identify the meaning of that
given shot or "gist of a scene" without remembering specific details.

@@

\subsection{Factor Graphs}
A \emph{factor graph}~\cite{kschischang2001factor} is a bipartite graph connecting two sets of nodes $X_G$ and $F_G$
representing random variables and factors.
Each factor is described by function $\phi$ dependent only on the variables $x_\phi$
to which the factor is connected.
Thus, a factor graph can be seen as a description of probability density function obtained
by a product of all the factors. In order to represent the probability,
a normalization factor needs to be introduced, resulting in the following equation:

>>>>

For example \glspl{SVM}, in their basic form, are only able to handle linear spaces.
But the classes are most of the time not linearly separable in the input
space. Although there might exist a transformation $\phi$ from the original
space into a space $H$ where the input becomes linearly separable.

@@

The most used method was presented by \cite{lowe1999object} and its based on
building a feature vector for each image. Each of those features is based on
\emph{interesting points} detected by the presence of maxima and minima of
difference of Gaussian functions applied in a scale-space.
The scale space is used to provide scale invariant detection. Gaussian functions
are used as they are the only way to model a linear scale-space.
Each interesting point is then described by a container that is rotation
invariant.

@@

\subsubsection*{Gist of a Scene}
\label{sec:gist}
Oliva and Torralba~\cite{oliva2006building} argue that fast scene recognition does not need to be
built on top of object recognition but can be analyzed by scene-centered
mechanisms.
They defend that position by pointing out behaviours on human vision:
when provided with a glance of a shot a person can identify the meaning of that
given shot or "gist of a scene" without remembering specific details.

@@

\subsection{Factor Graphs}
A \emph{factor graph}~\cite{kschischang2001factor} is a bipartite graph connecting two sets of nodes $X_G$ and $F_G$
representing random variables and factors.
Each factor is described by function $\phi$ dependent only on the variables $x_\phi$
to which the factor is connected.
Thus, a factor graph can be seen as a description of probability density function obtained
by a product of all the factors. In order to represent the probability,
a normalization factor $Z$ needs to be introduced, resulting in the following equation:

### 2011-07-04 thesis/semantic-mapping.tex
Some language fixes by LPR and Michael.

% motivation
Humans associate concepts to areas that relate to the functionality of those. That
association is either useful by explaining where to find certain objects as well to
understand the types of activities to perform on certain areas. 
Those human-concepts are meant to be meaningful for humans and by incorporating conceptual
knowledge about them it is expected to improve the agent ability to interact and communicate
with humans and to perform human-like tasks.

@@

% Importance of spatial and conceptual knowledge
In order to build such a semantic map, it is important to define the spatial and conceptual knowledge
that allows an agent to map between its machine-friendly representation of the world and the high-level semantics
characteristic of the human world.
That knowledge needs to define concepts and how to segment and identify them from the low-level representation,
it needs also to define how the concepts relate to each other and attribute meanings to them.
In the case of indoor environments this can be seen for example as: rooms are a unit of division of area and are
often separated by doors, each room can also be characterized by the objects inside, size and shape, and they
can be categorized in meaningful concepts such as: kitchen, library, corridor, etc\dots

@@

% here a here a method is used that does semantic mapping and represents conceptual knowledge
% on a probabilistic fashion.
This thesis motivation lies on developing methods for detecting knowledge gaps on the spatial
semantic knowledge of a mobile agent. For that spatial knowledge needs to be defined, as well
the methods used to perform the semantic mapping.
In that context this chapter presents \gls{Dora}, a system that represents its spatial knowledge
with the representation proposed by \cite{pronobis2010ias}, and performs semantic mapping using
inference on a probabilistic graphical model.

@@

\subsection{Uncertain Sensing and Implicit Factors}
\label{sec:cues-from-low-level}
In a realistic scenario, classification tasks are by their nature uncertain.
And although a trained classifier is able to decide which class is more likely, there is
interest in handling uncertainty produced by those to make decisions more robust.

>>>>

% motivation
Humans associate concepts to areas that relate to their functionality. That
association is either useful by explaining where to find certain objects as well to
understand the types of activities to perform on certain areas. 
Those human-concepts are meant to be meaningful for humans and by incorporating conceptual
knowledge about them it is expected to improve the agent ability to interact and communicate
with humans and to perform human-like tasks.

@@

% Importance of spatial and conceptual knowledge
In order to build such a semantic map, it is important to define the spatial and conceptual knowledge
that allows an agent to map between its machine-friendly representation of the world and the high-level semantics
characteristic of the human world.
That knowledge needs to define and describe how to segment and identify concepts from the low-level representation.
It also needs to define how the concepts relate to each other and attribute meanings to them.
In the case of indoor environments this can be seen for example as: rooms are a unit of division of area and are
often separated by doors, each room can also be characterized by the objects inside, size and shape, and they
can be categorized in meaningful concepts such as: kitchen, library, corridor, etc\dots

@@

% here a here a method is used that does semantic mapping and represents conceptual knowledge
% on a probabilistic fashion.
This thesis motivation lies on developing methods for detecting knowledge gaps on the spatial
semantic knowledge of a mobile agent. For that, spatial knowledge needs to be defined, as well
the methods used to perform the semantic mapping.
In that context this chapter presents \gls{Dora}, a system that represents its spatial knowledge
with the representation proposed by \cite{pronobis2010ias}, and performs semantic mapping using
inference on a probabilistic graphical model.

@@

\subsection{Uncertain Sensing and Implicit Factors}
\label{sec:cues-from-low-level}
In a realistic scenario, classification tasks are, by their nature, uncertain.
And although a trained classifier is able to decide which class is more likely, there is
interest in handling uncertainty produced by those to make decisions more robust.
